---
title: "Alzheimers Behaviorial Phenotypes"
output: html_document
date: "2024-07-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=10, warning=FALSE, message=FALSE)
```

**First we need some R libraries:**

```{r}
#remotes::install_github("https://github.com/monarch-initiative/monarchr", ref = "engines")

suppressMessages({
	library(monarchr)
	library(ggraph)
	library(stringr) # for str_wrap
	library(tidygraph) # activate() etc.
})
```


Let's start by looking for Alzheimer's. We'll use the search here in R to see what we can see:

```{r echo=TRUE}
monarch_engine() |>
	search_nodes("Alzheimer")
```

There it is, `MONDO:0004975`. Let's get it and all the descendants, and make a quick plot. We'll investigate the phenotypes connected to these below.


```{r echo=TRUE}
alz <- monarch_engine() |>
	fetch_nodes(query_ids = "MONDO:0004975") |> # Down
	expand(predicates = "biolink:subclass_of", direction = "in", transitive = TRUE)

alz

quickplot <- function(g) {
	ggraph(g, layout = "sugiyama") +
		geom_edge_link(aes(color = predicate),
									 arrow = arrow(length = unit(2, 'mm'), type = "closed"),
									 end_cap = circle(2.5, 'mm')) +
		geom_node_point(aes(color = pcategory),
										size = 3) +
		geom_node_label(aes(label = str_wrap(name, 20)),
										size = 2,
										repel = TRUE,
										check_overlap = TRUE,
										fill = "#FFFFFF88")	
}


quickplot(alz)
```

Cool. Let's summarize the neighborhood in this graph to see how many phenotypes etc. are connected.

```{r echo=TRUE}
alz |> summarize_neighborhood(summarize = "nodes")
```

Now, let's look for behavior-related entities.

```{r echo=TRUE}
monarch_engine() |>
	search_nodes("behavior")
```

"Atypical behavior" sounds likely. We'll get it's descendents and ancestors.

```{r echo=TRUE}
atypical_behaviors_desc <- monarch_engine() |>
	fetch_nodes(query_ids = "HP:0000708") |>
	expand(predicates = "biolink:subclass_of", direction = "in", transitive = TRUE)

atypical_behaviors_desc

atypical_behaviors_anc <- monarch_engine() |>
	fetch_nodes(query_ids = "HP:0000708") |>
	expand(predicates = "biolink:subclass_of", direction = "out", transitive = TRUE)

atypical_behaviors_anc
```

Over 700 subtypes of atypical behavior, and 26 ancestors; let's plot the latter to see if there are more general terms we should consider.


```{r echo=TRUE}
quickplot(atypical_behaviors_anc)

```

So there are a two higher-level entities that we might use instead of Atypical behavior - "abnormal behavior" and "Abnormality of mental function." Let's get the subtypes of both of those.


```{r echo=TRUE}
atypical_behaviors <- atypical_behaviors_anc |>
	activate(nodes) |>
	filter(name == "abnormal behavior") |>
	expand(predicates = "biolink:subclass_of", direction = "in", transitive = TRUE)
```

Next, the complete subtypes for "Abnormality of mental function".

```{r echo=TRUE}
atypical_mental_functions <- atypical_behaviors_anc |>
	activate(nodes) |>
	filter(name == "Abnormality of mental function") |>
	expand(predicates = "biolink:subclass_of", direction = "in", transitive = TRUE)

atypical_mental_functions
```

Let's filter the nodes in these graphs to see what they have in common and uniquely. We'll also add `type` labels to the result for later use, with `both` meaning a phenotype that is a descendant of both Behavioral and Mental abnormalities.

```{r}
# first: what's in atypical behaviors not in atypical mental functions
behavioral_only <- atypical_behaviors |>
	activate(nodes) |>
	filter(!id %in% nodes(atypical_mental_functions)$id) |>
	mutate(type = "behavioral")

behavioral_only


# what's in atypical mental functions not atypical behaviors
mental_only <- atypical_mental_functions |>
	activate(nodes) |>
	filter(!id %in% nodes(atypical_behaviors)$id) |>
	mutate(type = "mental")

mental_only

# and what they have in common
behavioral_and_mental <- atypical_behaviors |>
	activate(nodes) |>
	filter(id %in% nodes(atypical_mental_functions)$id) |>
	mutate(type = "both")

behavioral_and_mental
```
Ok, 24 behavioral only, 97 mental only, 724 both.

Let's join them back together (bringing in the labels we defined above). We'll also grab the phenotypes associated with Alzheimers (along with the nodes for Alzheimers itself), then join the two together to bring in the label info. We'll need to then cut this back down to just the nodes that are related to Alzheimers. Finally we'll create a plot of Alzheimers phenotypes labeled by their type.


```{r}
# combine the three sets above into a single table
all_behavioral_mental_phenos <- behavioral_only |>
	graph_join(mental_only) |>
	graph_join(behavioral_and_mental)

# alzheimers diseases and phenos
alz_phenos <- alz %>%
 	expand(result_categories = "biolink:PhenotypicFeature")

# all behavioral, mental, and alzheimers phenotypes, labeled
all_behavioral_mental_alz_phenos <- alz_phenos |>
	graph_join(all_behavioral_mental_phenos) |>
	mutate(type = ifelse(is.na(type), "other", type))

# finally, let's flush out the nodes that are not in the alz_phenos graph
alz_phenos_labeled <- all_behavioral_mental_alz_phenos |>
	activate(nodes) |>
	filter(id %in% nodes(alz_phenos)$id)

alz_phenos_labeled

ggraph(alz_phenos_labeled, layout = "sugiyama") +
		geom_edge_link(aes(color = predicate),
									 arrow = arrow(length = unit(2, 'mm'), type = "closed"),
									 end_cap = circle(2.5, 'mm')) +
		geom_node_point(aes(color = paste(pcategory, type)),
										size = 3) +
		geom_node_label(aes(label = str_wrap(name, 20)),
										size = 2,
										repel = TRUE,
										check_overlap = TRUE,
										fill = "#FFFFFF88")	
```

And here's a handy list of Alzheimers phenotypes, labeled with type as behavioral, mental, both, or other.

```{r echo=TRUE}
#ggraph(behavioral_mental_alz_phenos, layout = "sugiyama") +

alz_phenos_labeled |>
	activate(nodes) |>
	filter(pcategory == "biolink:PhenotypicFeature") |>
	select(id, name, type, description) |>
	arrange(type)
```



